---
title: "NYC Tree Census"
author: "Olachi Mbakwe and Arushi Singh"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

# Exploring the New York Tree Census

As the climate warms, there is more emphasis than ever on preserving and growing natural and natural areas. In a place like New York City, the most populated city in the United States, one would not expect this emphasis to affect change. However, New York City Parks Department, along with other private non-profit organizations have been incredibly successful in promoting the planting of new trees around the city, which benefits the community by providing greenspace and helps mitigate climate change by the increase of tree canopies. To track this change, we got datasets of the New York Tree Census, which occures every 10 years, from a website called NYCOpenData, where agencies can upload data for the public to view at their leisure. We focused on the over number/population of trees, their health, and the most popular species to track these changes.

```{r Data Wrangling, message=FALSE, warning=FALSE, include=FALSE}
#loading packages

library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(ggmap)
library(kableExtra)
library(janitor)
library(trackdown)

##Bringing in the newyork tree data sets for every 10 years 
nytrees_1995 <- read_csv("~/Desktop/1995_Street_Tree_Census.csv")
nytrees_2005 <- read_csv("~/Desktop/2005_Street_Tree_Census.csv")
nytrees_2015 <- read_csv("~/Desktop/2015_Street_Tree_Census_-_Tree_Data.csv")
#nytrees_species <- read_csv("Downloads/archive/new_york_tree_species.csv")

#Data wrangling

nytrees_1995 <- nytrees_1995 %>%
  select(Spc_Common,Condition,Borough,Longitude,Latitude,Zip_New)

nytrees_2005 <- nytrees_2005 %>%
  select(spc_common,status,boroname,latitude,longitude,zipcode)

nytrees_2015 <- nytrees_2015 %>%
  select(spc_common,health,borough,latitude,longitude,postcode)

```

# Exploring Population Density and Growth, alongside Tree Health

To start our analysis, we wanted a visualization of the most basic indicator of population growth over the years: the number of trees.

```{r Population Density, echo=FALSE, out.width="130%", fig.show='hold',message=FALSE, warning=FALSE}
## creating map Visualization
##using ggmaps 

newyork.map <- get_map(location = 'New York', maptype= "hybrid",source='google',zoom=11)

#1995
ggmap(newyork.map) + geom_point(data = nytrees_1995, 
aes(x=Longitude,y=Latitude,colour = Condition),size=.05)+
  scale_color_manual(
    values = c(Dead = "#282626",
               Excellent = "#16DE0C",
               Good = "#3D6C08",
               Poor = "#E50E0E")
  ) +
  labs(
    title = "New York Tree Census 1995 map"
  ) + xlab("Longitude") + ylab("Latitude") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold",
                              hjust = 0.5),
  )

#2005
ggmap(newyork.map) + geom_point(data = nytrees_2005, aes(x=longitude,y=latitude,colour = status),size=.05)+
  scale_color_manual(
    values = c(Dead = "#282626",
               Excellent = "#16DE0C",
               Good = "#3D6C08",
               Poor = "#E50E0E")
  ) +
  labs(
    title = "New York Tree Census 2005 map"
  ) +
  xlab("Longitude") + ylab("Latitude") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold",
                              hjust = 0.5),
  )

#2015
ggmap(newyork.map) + 
  geom_point(data = nytrees_2015, aes(x=longitude,y=latitude,colour = health),size=.05)+
  scale_color_manual(
    values = c(Fair = "#16DE0C",
               Good = "#3D6C08",
               Poor = "#E50E0E")  
  ) +
  labs(
    title = "New York Tree Census 2015 map",
    color = "Condition of the Tree"
  ) + xlab("Longitude") + ylab("Latitude") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold",
                              hjust = 0.5),
    )
```


The figures above show that from 1995 to 2015, the number of trees have only increased. Upon further inspection, you can see that these increases are initially mostly in area, but by 2015 the trees are also incredibly dense. This dramatic increase is still continuing today, as an effort from the New York Parks department, as well as Trees for Public Health to help mitigate the effects of climate change on the city, and is arguably the cheapest way to do so. By planting more trees, New York City's Tree Canopy is growing. A tree canopy, by the simplest definition are the cover provided by trees. Tree canopies help mitigate climate change because they store CO2, resulting in less of the gas contributing to the greenhouse gases effect. Tree canopies also reduce energy use by cooling neighborhoods by up to 10 degrees, without reducing energy needs. All in all, this increase in tree population has only been benefiting the people of New York City.

To go further, we decided to track the health alongside the population growth.

```{r Comparing Health 1995, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE, out.width="50%"}
###Making Two separate graph for good and bad condition

#1995
#Good Conditions
new_nytrees_1995_good <- nytrees_1995 %>% select(Longitude,Latitude,Condition) %>% filter(Condition %in% c("Good", "Excellent"))
ggmap(newyork.map) + geom_point(data = new_nytrees_1995_good, aes(x=Longitude,y=Latitude,colour = Condition),size=.05)+
  scale_color_manual(
    values = c(Excellent = "#16DE0C",
               Good = "#3D6C08")
  ) +
  labs(
    title = "Good Health",
    color = "Condition of the Tree"
  ) +
  xlab("Longitude") + ylab("Latitude") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
  )

#Bad Conditions
new_nytrees_1995_bad <- nytrees_1995%>% select(Longitude,Latitude,Condition) %>% filter(Condition %in% c("Dead", "Poor"))
ggmap(newyork.map) + geom_point(data = new_nytrees_1995_bad, aes(x=Longitude,y=Latitude,colour = Condition),size=.05)+
  scale_color_manual(
    values = c(Dead = "#282626",
               Poor = "#E50E0E")
  ) +
  labs(
    title = "Bad Health", 
    color = "Condition of the Tree"
  ) +
  xlab("Longitude") + ylab("Latitude") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
  )
```

The figure above displays the location of all of the trees in the data set, but also notes the health of each tree. As we compare the two graphs, we can see that the trees that are dead or in poor health are in heavily industrialized areas or in less populated areas. This leads me to infer that these trees simply aren't receiving the care that they need to thrive. The trees in excellent and good health look like they are concentrated in parks, as well as being intentionally planted in the city.

```{r Comparing Health 2005, echo=FALSE, fig.show='hold', out.width= "50%", message=FALSE, warning=FALSE}


#2005
#Good Conditions
new_nytrees_2005_good <- nytrees_2005%>% select(longitude,latitude,status) %>% filter(status %in% c("Good", "Excellent"))
ggmap(newyork.map) + geom_point(data = new_nytrees_2005_good, aes(x=longitude,y=latitude,colour = status),size=.05)+
  scale_color_manual(
    values = c(Excellent = "#16DE0C",
               Good = "#3D6C08")
  ) +
  labs(
    title = "Good Health",
    color = "Condition of the Tree"
  ) +
  xlab("Longitude") + ylab("Latitude") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
  )

#Bad Conditions
new_nytrees_2005_bad <- nytrees_2005%>% select(longitude,latitude,status) %>% filter(status %in% c("Dead", "Poor"))
ggmap(newyork.map) + geom_point(data = new_nytrees_2005_bad, aes(x=longitude,y=latitude,colour = status),size=.05)+
  scale_color_manual(
    values = c(Dead = "#282626",
               Poor = "#E50E0E")
  ) +
  labs(
    title = "Bad Health", 
    color = "Condition of the Tree"
  ) +
  xlab("Longitude") + ylab("Latitude") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
  )
```

In 2005, we can see that despite the seemingly good news that the tree population had increased, many of those trees were in poor health or had already died. Some good news from this visualization is that the excellent to good ratio in the first visualization is much improved. In conclusion, in 2005, the Parks department still had a ways to go to improve their tree maintenance.

```{r Comparing Health 2015, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE, out.width="50%"}

#2015
#Good Conditions
new_nytrees_2015_good <- nytrees_2015 %>% select(longitude,latitude,health) %>% filter(health %in% c("Good", "Fair"))
ggmap(newyork.map) + geom_point(data = new_nytrees_2015_good, aes(x=longitude,y=latitude,colour = health),size=.05)+
  scale_color_manual(
    values = c(Good = "#3D6C08",
               Fair = "#16DE0C")
  ) +
  labs(
    title = "Good Health",
    color = "Condition of the Tree"
  ) +
  xlab("Longitude") + ylab("Latitude") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
  )

#Bad Conditions
new_nytrees_2015_bad <- nytrees_2015 %>% select(longitude,latitude,health) %>% filter(health %in% c("Poor"))
ggmap(newyork.map) + geom_point(data = new_nytrees_2015_bad, aes(x=longitude,y=latitude,colour = health),size=.05)+
  scale_color_manual(
    values = c(Poor = "#E50E0E")
  ) +
  labs(
    title = "Bad Health", 
    color = "Condition of the Tree"
  ) +
  xlab("Longitude") + ylab("Latitude") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
  )
```

By 2015, the health of trees had had much improvement and the majority of trees appear to be in good health! The ratio of good: poor trees is much improved and the overall health of the trees in the city have grown, while also maintaining the population growth rate.

As we can see from this set of visualizations, from 1995 to 2015, not only has the number of trees increased in that time frame, the overall health of them has as well. Given the city-environment, I initally expected this trend to be the opposite. However, it is incredibly motivating to see how the effort put it by the Parks department, and other New York advocacy groups are paying off and providing some protection for New Yorkers against the growing climate crisis.

# Trees by Borough

```{r Tables- Popularity, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE, out.width="33%"}
##Borough has the most tree
#Total Number of trees in each Borough

Borough_1995 <- data.frame(sort(table(nytrees_1995$Borough), decreasing = TRUE))
Borough_2005 <- data.frame(sort(table(nytrees_2005$boroname), decreasing = TRUE))
Borough_2015 <- data.frame(sort(table(nytrees_2015$borough), decreasing = TRUE))


# Use the colnames function to specify the column names
colnames(Borough_1995) <- c("Borough", "Total Number of Trees")
colnames(Borough_2005) <- c("Borough", "Total Number")
colnames(Borough_2015) <- c("Borough", "Total Number")


# View the resulting table
#1995
kbl(head(Borough_1995, 5), caption = "Total Number of Trees in each Borough - 1995" )%>%
  kable_styling(bootstrap_options = "bordered")%>%
  kable_paper(full_width = F)

#2005
kbl(head(Borough_2005, 5), caption = "Total Number of Trees in each Borough - 2005" )%>%
  kable_styling(bootstrap_options = "bordered")%>%
  kable_paper(full_width = F)

#2015
kbl(head(Borough_2015, 5), caption = "Total Number of Trees in each Borough - 2015" )%>%
  kable_styling(bootstrap_options = "bordered")%>%
  kable_paper(full_width = F)
```

The tables above display the number of trees per borough. Although the number of trees has grown, Manhattan consistenly has the lowest tree population out of all of the boroughs. This makes sense, given that Manhattan has the smallest area out of all of the boroughs and is also the most industrialized. Manhattan has an area of 23 square miles and is home to almost 1.7 million people. Put in this context, it makes sense that it has the least number of trees. Queens consistently has had the highest tree population over the year, with Brooklyn at second place and a difference of about 10,000 trees between them each year. Queens has an area of 109 square miles, while Brooklyn has 42 square miles. They both also have about the same population, so given the difference in area, yet approximate equality in population, it makes sense why Queens is home to so many more trees than the other boroughs. 

# Tree Health Per Borough

To explore trees by borough more thoroughly, we decided to look at the health of trees by borough more directly. 

```{r Bar Charts- Popularity 1995, echo=FALSE, fig.width=15, message=FALSE, warning=FALSE}
#Using Facet Visualization on the Health/Condition of the tree per borough
##1995
nytrees_1995 %>%
  filter(Condition %in% c("Good", "Excellent", "Poor", "Dead")) %>%
  ggplot() +
  aes(x = Borough, fill = Borough) +
  geom_bar() +
  scale_fill_manual(
    values = c(Bronx = "#814198",
               Brooklyn = "#0370B7",
               Manhattan = "#E59C0C",
               Queens = "#E01D1D",
               `Staten Island` = "#4B935B")
  ) +
  labs(
    x = "Borough",
    y = "Total num of Tree",
    title = "1995 Trees Condition per Borough",
    fill = "Boroughs"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold",
                              hjust = 0.5),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  ) +
  facet_wrap(vars(Condition), scales = "free")
```

In 1995, we can see that Queens had the highest amount of dead trees, as well as one of the highest amounts of trees in excellent condition. Staten Island has the best proportion of trees in excellent condition to dead trees. 

```{r Bar Charts- Popularity 2005, echo=FALSE, fig.width=15, message=FALSE, warning=FALSE}
#2005
ggplot(nytrees_2005) +
  aes(x = boroname, fill = boroname) +
  geom_bar() +
  scale_fill_manual(
    values = c(Bronx = "#814198",
               Brooklyn = "#0370B7",
               Manhattan = "#E59C0C",
               Queens = "#E01D1D",
               "5" = "#4B935B")
  ) +
  labs(
    x = "Borough",
    y = "Total num of Tree",
    title = "2005 Trees Condition per Borough"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold",
                              hjust = 0.5),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  ) +
  facet_wrap(vars(status), scales = "free")
```

```{r Bar Charts- Popularity 2015, echo=FALSE, fig.width=15, message=FALSE, warning=FALSE}
#2015
nytrees_2015 %>%
  filter(health %in% c("Good", "Fair", "Poor")) %>%
  ggplot() +
  aes(x = borough, fill = borough) +
  geom_bar() +
  scale_fill_manual(
    values = c(Bronx = "#814198",
               Brooklyn = "#0370B7",
               Manhattan = "#E59C0C",
               Queens = "#E01D1D",
               `Staten Island` = "#4B935B")
  ) +
  labs(
    x = "Borough",
    y = "Total num of Tree",
    title = "2015 Trees Condition per Borough"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15L,
                              face = "bold",
                              hjust = 0.5),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  )+
  facet_wrap(vars(health), scales = "free")


```

# Specifying Population Growth and Changes in Health by Tree Species

After looking at changes in the health of the trees, we decided to subset the data and just focus on the most popular species and analyze how the species themselves and their health changes over the years. 

```{r Most Popular Species Tables, echo=FALSE, fig.show='hold', message=FALSE, warning=FALSE, out.width="33%"}
#most popular tree species
mostPopular_1995 <- data.frame(sort(table(nytrees_1995$Spc_Common), decreasing = TRUE))
mostPopular_2005 <- data.frame(sort(table(nytrees_2005$spc_common), decreasing = TRUE))
mostPopular_2015 <- data.frame(sort(table(nytrees_2015$spc_common), decreasing = TRUE))


# Use the colnames function to specify the column names
colnames(mostPopular_1995) <- c("Tree Species", "Total Number")
colnames(mostPopular_2005) <- c("Tree Species", "Total Number")
colnames(mostPopular_2015) <- c("Tree Species", "Total Number")

# View the resulting table 
#1995
kbl(head(mostPopular_1995, 5), caption = "Top 5 common tree Species in 1995" )%>%
  kable_styling(bootstrap_options = "bordered")%>%
  kable_paper(full_width = F)

#2005
kbl(head(mostPopular_2005, 5), caption = "Top 5 common tree Species in 2005" )%>%
  kable_styling(bootstrap_options = "bordered")%>%
  kable_paper(full_width = F)

#2015
kbl(head(mostPopular_2015, 5), caption = "Top 5 common tree Species in 2015" )%>%
  kable_styling(bootstrap_options = "bordered")%>%
  kable_paper(full_width = F)

```
* i might need to cite the websites
From these tables, we can see that the most popular trees generally remained the same from 1995 to 2015. In 1995, we can see that the Norway Maple was the most popular species by a large margin. This, however, was problematic because Norway Maple Trees are an invasive species native to Continental Europe, Southern Norway, and Western Asia. Invasive species are any organism that are brought into a non-native environment and harm this new environment. Norway Maple Trees aid in declining biodiversity and make it harder for native species to make it to maturity. Callery Pear Trees are also invasive species, native to East Asia, specifically China. The London Planetree is a hybrid between an American Sycamore and a Non-Native tree. The Pin Oak and HoneyLocust are both species native to North America. Knowing this, it is incomprehensibly good that the invasive species are slowly becoming less popular, and the native tree species are becoming more popular. I was unable to find if this is an a focused effort on the Parks' Department part, but either way it is amazing news for the native New York species who rely on these native trees. The Norway Maple's decline in count in particularly awe-inspiring, considering it was the most popular species in 1995, and had expereinced a loss of 75,000 trees by 2015. 


The following visualizations depict the most popular species in each borough. 

```{r Tree Popularity by Borough 1995, echo=FALSE, fig.width=25, message=FALSE, warning=FALSE}


#Visualization using Bar plot on the most common species per borough

#1995
species_common_1995 <- nytrees_1995 %>%
  select(Spc_Common, Borough)%>%
  group_by(Spc_Common, Borough) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

trees_by_borough_1995 <- bind_rows(
  head(filter(species_common_1995, Borough == "Staten Island"),5),
  head(filter(species_common_1995, Borough == "Queens"),5),
  head(filter(species_common_1995, Borough == "Manhattan"),5),
  head(filter(species_common_1995, Borough == "Brooklyn"),5),
  head(filter(species_common_1995, Borough == "Bronx"),5)
)

ggplot(trees_by_borough_1995) +
  aes(x = Spc_Common, y = count, fill = Borough) +
  geom_col()+
    scale_fill_manual(
    values = c(Bronx = "#814198",
               Brooklyn = "#0370B7",
               Manhattan = "#E59C0C",
               Queens = "#E01D1D",
               `Staten Island` = "#4B935B")
  ) +
  labs(
    x = "Tree Species",
    y = "Count",
    title = "Top Five Most Common Species per Borough - 1995"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18L,
                              face = "bold",
                              hjust = 0.5),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  ) +
  facet_wrap(vars(Borough), scales = "free")
```

In 1995, we can see that the Maple Norway was most common in the Bronx, Queens, and Staten Island, while the London Planetree was most common in Brooklyn and the Honeylocust was most common in Manhattan. I think its also interesting how many species of Maple Trees are popular in these boroughs. Maple, Linden, and Gingko trees are all invasive species, but are considered New York City staples, especially in the fall. 

```{r Tree Popularity by Borough 2005, echo=FALSE, fig.width=25, message=FALSE, warning=FALSE}
#2005
species_common_2005 <- nytrees_2005 %>%
  select(spc_common, boroname)%>%
  group_by(spc_common, boroname) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

trees_by_borough_2005 <- bind_rows(
  head(filter(species_common_2005, boroname == "5"),5),
  head(filter(species_common_2005, boroname == "Queens"),5),
  head(filter(species_common_2005, boroname == "Manhattan"),5),
  head(filter(species_common_2005, boroname == "Brooklyn"),5),
  head(filter(species_common_2005, boroname == "Bronx"),5)
)

ggplot(trees_by_borough_2005) +
  aes(x = spc_common, y = count, fill = boroname) +
  geom_col() +
  scale_fill_manual(
    values = c(Bronx = "#814198",
               Brooklyn = "#0370B7",
               Manhattan = "#E59C0C",
               Queens = "#E01D1D",
               "5" = "#4B935B")
  ) +
  labs(
    x = "Tree Species",
    y = "Count",
    title = "Top Five Most Common Species per Borough - 2005"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18L,
                              face = "bold",
                              hjust = 0.5),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  ) +
  facet_wrap(vars(boroname), scales = "free")
```

In 2005, we see that most of the most popular trees stayed the same, with the exclusion of Silver Maples in Staten Island and Queens, Sugar Maples in the Bronxand Little Leaf Lindens in Brooklyn. We can also see the increase of native species in Queens and the decrease of Norway Maples in Staten Island. 

```{r Tree Popularity by Borough 2015, echo=FALSE, fig.width=25, message=FALSE, warning=FALSE}
#2015
species_common_2015 <- nytrees_2015 %>%
  select(spc_common, borough)%>%
  group_by(spc_common, borough) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

trees_by_borough_2015 <- bind_rows(
  head(filter(species_common_2015, borough == "Staten Island"),5),
  head(filter(species_common_2015, borough == "Queens"),5),
  head(filter(species_common_2015, borough == "Manhattan"),5),
  head(filter(species_common_2015, borough == "Brooklyn"),5),
  head(filter(species_common_2015, borough == "Bronx"),5)
)

ggplot(trees_by_borough_2015) +
  aes(x = spc_common, y = count, fill = borough) +
  geom_col() +
  scale_fill_manual(
    values = c(Bronx = "#814198",
               Brooklyn = "#0370B7",
               Manhattan = "#E59C0C",
               Queens = "#E01D1D",
               `Staten Island` = "#4B935B")
  ) +
  labs(
    x = "Tree Species",
    y = "Count",
    title = "Top Five Most Common Species per Borough - 2015"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 18L,
                              face = "bold",
                              hjust = 0.5),
    axis.title.y = element_text(face = "bold"),
    axis.title.x = element_text(face = "bold")
  ) +
  facet_wrap(vars(borough), scales = "free")
```

By 2015, native species are much more common in each borough. There are some new non native species, such as the Japanese Zeikova and the Sophora Japonica, both of which are species native to Japan. These species are both invasive, but have thrived in New York City due to their ability to grow well in tight spaces. The Honeylocust or London Planetree are the most popular in each borough, which is also reflected by the overall New York City counts. 

After analyzing the species by borough, we decided to look at which species are actually thriving, health wise.

```{r Tree Condition 1995, echo=FALSE, fig.width=15, message=FALSE, warning=FALSE}
##The most common species given the Health Condition 
##Creating a Bar plot Visualisation
#1995

Nytrees_1995 <- nytrees_1995 %>%
  select(Spc_Common, Condition)%>%
  group_by(Spc_Common, Condition) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

trees_by_condition_1995 <- bind_rows(
  head(filter(Nytrees_1995, Condition == "Excellent"),5),
  head(filter(Nytrees_1995, Condition == "Good"),5),
  head(filter(Nytrees_1995, Condition == "Poor"),5),
  head(filter(Nytrees_1995, Condition == "Dead"),5)
)

##Visualization 
ggplot(trees_by_condition_1995) +
  aes(x = Spc_Common, y = count) +
  geom_col(fill = "#053E05") +
  labs(
    x = "Tree Species ",
    y = "Count",
    title = "Top Five Tree Species given their Health Condition - 1995"
  ) +
  theme_minimal() +
  facet_wrap(vars(Condition), scales = "free")
```

In 1995, most of the dead trees were unfortunately unknown or unidentifiable. Predictably, the trees in excellent health were mostly the Norway Maple and the Callery Pear, which makes sense givern their invasive status. However, the Norway Maple had about the same number of trees in a dead or poor health status as there were trees in excellent status, indicating its future decline. Although there were less native trees at this point, most of them were in good or excellent health. 

```{r Tree Condition 2005, echo=FALSE, fig.width=15, message=FALSE, warning=FALSE}
#2005

Nytrees_2005 <- nytrees_2005 %>%
  select(spc_common, status)%>%
  group_by(spc_common, status) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

trees_by_condition_2005 <- bind_rows(
  head(filter(Nytrees_2005, status == "Excellent"),5),
  head(filter(Nytrees_2005, status == "Good"),5),
  head(filter(Nytrees_2005, status == "Poor"),5),
  head(filter(Nytrees_2005, status == "Dead"),5)
)

##Visualization
ggplot(trees_by_condition_2005) +
  aes(x = spc_common, y = count) +
  geom_col(fill = "#053E05") +
  labs(
    x = "Tree Species ",
    y = "Count",
    title = "Top Five Tree Species given their Health Condition - 2005"
  ) +
  theme_minimal() +
  facet_wrap(vars(status), scales = "free")
```

In 2005, most dead trees were still unknown species, while the most popular species in excellent health was the Callery Pear. Norway Maples were the most popular species in poor health, which makes sense as we consider its dramatic decline in population. The London Planetree was the second popular tree in excellent health and the species with the highest count of trees in good condition, which makes sense along with its small increase in population between 1995 and 2005. 

```{r Tree Condition 2015, echo=FALSE, fig.width=17, message=FALSE, warning=FALSE}
#2015 

Nytrees_2015 <- nytrees_2015 %>%
  select(spc_common, health)%>%
  group_by(spc_common, health) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

trees_by_condition_2015 <- bind_rows(
  head(filter(Nytrees_2015, health == "Excellent"),5),
  head(filter(Nytrees_2015, health == "Good"),5),
  head(filter(Nytrees_2015, health == "Fair"),5),
  head(filter(Nytrees_2015, health == "Poor"),5)
)

##Visualization 
ggplot(trees_by_condition_2015) +
  aes(x = spc_common, y = count) +
  geom_col(fill = "#053E05") +
  labs(
    x = "Tree Species ",
    y = "Count",
    title = "Top Five Tree Species given their Health Condition - 2015"
  ) +
  theme_minimal() +
  facet_wrap(vars(health), scales = "free")
```

In 2015, instead of recording Dead, Poor, Fair, and Excellent status, the census changed to recording just Poor, Fair, and Good. Because of this, it is a little difficult to direcly compare changes without making some inferences. Nevertheless, here we see that Norway Maple had the most trees in poor health, along with a lower count overall. The species with the most trees in good and fair health was the London Planetree. I was surprised to see that one of the most popular species in good health is the Japanese Zelkova. This goes to show how quickly invasive species and dominante existing thriving species.

# Exploring the most Popular Species further

Although the most popular species in each borough differs, we decided to go back to the most popular species in the city overall and track their health through the years by borough. Since the popular species overral didn't change, we were able to compare the changes easily. 

```{r Most Popular and Health 1995, echo=FALSE, fig.width=11, message=FALSE, warning=FALSE}

mostPop1995 <- nytrees_1995 %>%
    group_by(Spc_Common) %>%
    summarise(count = n()) %>%
    top_n(n = 5, wt = count)

mostPopData1995 <- filter(nytrees_1995, Spc_Common == c("HONEYLOCUST","LONDON PLANETREE","MAPLE, NORWAY", "OAK, PIN","PEAR, CALLERY"))

mostPopFinal1995 <- mostPopData1995 %>%
    group_by(Spc_Common, Borough, Condition) %>%
    summarise(count = n()) %>%
    top_n(n = 5, wt = count)

ggplot(
  data = mostPopFinal1995, 
  aes(x = Condition, y = count, fill = Spc_Common)) +
  geom_col() +
  facet_wrap(.~Borough, scales = "free") +
  theme_minimal() + 
  xlab("Condition of Tree (Scale of Dead to Excellent, including Unknown)")+
  ylab("Count of Trees") +
  ggtitle("The Health of the Most Popular Tree Species in 1995 by Borough")+
  labs(fill = "5 Most Popular Species")
  
```

In 1995, the ratio of species in good health in the Bronx and Staten Island is pretty even. However, in Brooklyn, Manhattan and Queens there is one species that is clearly in better health compared to the others. In Brooklyn, thats the London Planetree, in Manhattan its the Honeylocust, and in Queens it's the Norway Maple. The tree most relatively in poor health is the Norway Maple. 

```{r Most Popular and Health 2005, echo=FALSE, fig.width=10, message=FALSE}

mostPop2005 <- nytrees_2005 %>%
    group_by(spc_common) %>%
    summarise(count = n()) %>%
    top_n(n = 5, wt = count)

mostPopData2005 <- filter(nytrees_2005, spc_common == c("HONEYLOCUST","LONDON PLANETREE","MAPLE, NORWAY", "OAK, PIN","PEAR, CALLERY"))

mostPopFinal2005 <- mostPopData2005 %>%
    group_by(spc_common, boroname, status) %>%
    summarise(count = n()) %>%
    top_n(n = 5, wt = count)

ggplot(
  data = mostPopFinal2005, 
  aes(x = status, y = count, fill = spc_common)) +
  geom_col() +
  facet_wrap(.~boroname, scales = "free") +
  theme_minimal() + 
  xlab("Condition of Tree (Scale of Dead to Excellent, including Unknown)")+
  ylab("Count of Trees") +
  ggtitle("The Health of the Most Popular Tree Species in 2005 by Borough")+
  labs(fill = "5 Most Popular Species")
  
```

In 2005, only Bronx has relatively even ratios of species in excellent health. In Staten Island, the Callery Pear is the healthiest tree. In Brooklyn it's the London Planetree, and in Manhattan its the Honeylocust. In Queens, the Norway Maple is the healthiest by a much smaller margin than in 1995. In Queens, the Norway Maple is also the one in the most poor health, but most of the other boroughs have less dramatic ratios of species in poor health.

```{r Most Popular and Health 2015, echo=FALSE, fig.width=10, message=FALSE, warning=FALSE}

mostPop2015 <- nytrees_2015 %>%
    group_by(spc_common) %>%
    summarise(count = n()) %>%
    top_n(n = 5, wt = count)

mostPopData2015 <- filter(nytrees_2015, spc_common == c("Callery pear","honeylocust","London planetree", "Norway maple","pin oak"))

mostPopFinal2015 <- mostPopData2015 %>%
    group_by(spc_common, borough, health) %>%
    summarise(count = n()) %>%
    top_n(n = 5, wt = count)

ggplot(
  data = mostPopFinal2015, 
  aes(x = health, y = count, fill = spc_common)) +
  geom_col() +
  facet_wrap(.~borough, scales = "free") +
  theme_minimal() + 
  xlab("Condition of Tree (Scale of Dead to Excellent, including Unknown)")+
  ylab("Count of Trees") +
  ggtitle("The Health of the Most Popular Tree Species in 2015 by Borough")+
  labs(fill = "5 Most Popular Species")
  
```

By 2015, the ratios of species in fair and poor health is relatively even once again. However, the species in good health differs wildly by each borough. In Brooklyn and Queens, the London Planetree is the most popular tree in good health, while in Manhattan and the Bronx, it's the Honeylocust. In Staten Island, its the Callery Pear tree. This analysis goes to show that despite increases or decreases in species population, the trees themselves may be in poor health, which can be a predictor of a future decline in species popularity. This also applies in the opposite way, where if the ratio of trees in good or excellent health to poor health is favoring the trees is good or excellent health, there will likely be an increase in that tree species population. 

```{r Trackdown, eval=FALSE, include=FALSE}

trackdown::upload_file(
file = "TrackdownDemo.Rmd",
hide_code = TRUE,
rich_text = TRUE,
path_output = "TrackdownDemo.pdf"
)
trackdown::update_file(
file = "TrackdownDemo.Rmd",
hide_code = TRUE,
rich_text = TRUE,
path_output = "TrackdownDemo.pdf"
)
trackdown::download_file(
file = "TrackdownDemo.Rmd"
)
trackdown::render_file(
file = "TrackdownDemo.Rmd"
```

# Final Thoughts

In conclusion, New York City has absolutely transformed in terms of tree species and health overall and by borough over the years. It's been incredibly interesting to research how different things have effected tree health and how invasive species can be incredibly detrimental to native species' health. Overall, I think the efforts put in by the city's Parks department and other advocacy groups and agencies are wildly paying off, and hopefully will end up mitigating the effects of climate change in New York City. 